---
title: "Magic in R and Rust"
output:
  html_document:
    self_contained: true
    mathjax: default
date: "2025-05-06"
always_allow_html: true
editor_options:
  chunk_output_type: console

---
<!-- load MathJax from CDN -->
<script src="https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-chtml.js"></script>


## Overview of Magic implemented in R and Rust

*(“Markov Affinity-based Graph Imputation of Cells” – van Dijk et al., 2018)*

| Symbol       | Size    | Meaning                                                                       |
| ------------ | ------- | ----------------------------------------------------------------------------- |
| **X**        | *n × g* | raw (or PCA-reduced) expression; rows = cells, cols = genes                   |
| **d(i,j)**   | –       | Euclidean (or cosine) distance between cell *i* and *j* in PCA space          |
| **σᵢ**       | –       | **local scale** for cell *i* (distance to its *k*-th nearest neighbour)       |
| **A**        | *n × n* | symmetrised, locally scaled affinity (“heat”) matrix                          |
| **D**        | *n × n* | diagonal degree matrix, **D**<sub>ii</sub> = ∑<sub>j</sub> **A**<sub>ij</sub> |
| **P = D⁻¹A** | *n × n* | row-stochastic Markov transition matrix                                       |
| **t**        | –       | diffusion time (walk length)                                                  |
| **Ŷ**        | *n × g* | imputed expression after diffusion                                            |

---

#### 1 k-NN graph in latent space

Compute *k* nearest neighbours in the *m*-dimensional PCA space
(typically *k* ≈ 30, *m* ≈ 100).

This is done by Seurat or related packages.

#### 2 Adaptive Gaussian kernel

For each edge (*i*, *j*) in that graph set

$$
A_{ij} \;=\; 
\begin{cases}
  \displaystyle
  \exp\!\Bigl(-\frac{d(i,j)^2}{\sigma_i\,\sigma_j}\Bigr),
    & j\in\mathrm{kNN}(i),\\[6pt]
  0, & \text{otherwise.}
\end{cases}
$$



Local bandwidths σᵢ make the kernel **anisotropic** so dense and sparse
regions of the manifold are treated equally.

##### Symmetrise

$$
A \;←\; \tfrac12\,(A + A^{\!\top})
$$

#### 3 Markov normalisation

$$
P \;=\; D^{-1}A,
\quad
D_{ii} = \sum_{j} A_{ij}.
$$

Now each row of **P** sums to 1 → one step of a random walk on the cell graph.

This is done in R:

```{r, eval=F}
row_sums <- Matrix::rowSums(A)
P <- Matrix::Diagonal(x = 1 / pmax(row_sums, 1e-12)) %*% A
```

#### 4 Diffusion (raise to power *t*)

There are two equivalent views:

* **Spectral**
  Diagonalise **P** = Ψ Λ Ψ⁻¹ with eigenvalues 1 = λ₀ > λ₁ ≥ λ₂ …
  Then

  $$
  P^{t} \;=\; \Psi\,\Lambda^{t}\,\Psi^{-1},
  \qquad
  \Lambda^{t} = \operatorname{diag}(\lambda_k^{t}).
  $$

  Small eigenmodes (high-frequency noise) decay as λₖᵗ.

* **Random-walk**
  Entry (i,j) of **Pᵗ** is the probability that a random walk of
  length *t* starting at cell *i* ends in cell *j*.

#### 5 Imputation by heat propagation

Apply the diffusion operator to every gene vector:

$$
\boxed{\;
\hat X = P^{t}\,X
\;}
\quad\Longrightarrow\quad
\hat x_{ig} = \sum_{j=1}^{n} (P^{t})_{ij}\; x_{jg}.
$$

Each imputed expression value becomes a weighted average over the
*t*-step neighbourhood, smoothing drop-outs while respecting manifold
structure.

---

### Relation to the continuous heat equation

On a smooth data manifold **M**, the generator **L** = *I* – **P**
approximates the Laplace–Beltrami operator Δ<sub>M</sub>.
Diffusion time *t* therefore controls how far one solves the
heat-equation

$$
\frac{\partial u}{\partial t} \;=\; -\Delta_M\,u, \qquad u(0)=X.
$$

MAGIC performs this on a *graph* and stops at a user-chosen *t*
(often 2–6); too small → little denoising, too large → over-smoothing.


---

## Blended MAGIC imputation

We have added an additional lever which controls the strength of the diffusion process.

Let

* $X\in\mathbb{R}^{n\times g}$ be the original (cells × genes) expression matrix,
* $P\in\mathbb{R}^{n\times n}$ the row-stochastic diffusion operator,
* $t\in\mathbb{N}$ the number of diffusion steps,
* $\widetilde X = P^{\,t} X$ the fully-diffused matrix,
* $\alpha\in[0,1]$ the blending weight.

We define the **blended imputation**

$$
X_{\rm imp}(\alpha)
\;=\;
(1 - \alpha)\,X
\;+\;
\alpha\,\widetilde X
\;=\;
(1 - \alpha)\,X
\;+\;
\alpha\,P^{\,t}\,X
\;\in\mathbb{R}^{n\times g}.
$$

### Special cases

\[
\begin{aligned}
\alpha = 0:\quad &X_{\rm imp}(0) = X,\\
\alpha = 1:\quad &X_{\rm imp}(1) = P^{t}X,\\
0 < \alpha < 1:\quad &\text{Partial smoothing (convex blend of $X$ and $P^{t}X$).}
\end{aligned}
\]


### Per-entry formula

For each cell $i$ and gene $j$:

$$
X_{\rm imp}^{(i,j)}(\alpha)
\;=\;(1 - \alpha)\,X^{(i,j)}
\;+\;\alpha\sum_{k=1}^{n} \bigl[P^{\,t}\bigr]_{ik}\,X^{(k,j)}.
$$

### Interpretation

* **$\alpha$** directly controls the **strength** of smoothing:

  * Small $\alpha$ preserves more of the raw, per-cell expression.
  * Large $\alpha$ favors neighborhood-averaged values.
* Choosing $\alpha$ between 0 and 1 lets you trade off **noise reduction** vs.
  **signal fidelity** without changing $t$ or the underlying graph.

---



## How to run MAGIC in R/Rust

First we take some data, subset to the tumor cells that express modest levels of CD33.
```{r}
suppressPackageStartupMessages({
  library(Seurat)
  library(scCustomize)
  library(rustytools)
  library(magrittr)
})


seu <- readRDS("~/Fred Hutch Cancer Center/Furlan_Lab - General/experiments/patient_marrows/annon/AML101/aml101.cds")
seu$sb <- seu$geno %in% "0" & seu$seurat_clusters %in% c("0", "1", "2", "3", "4", "11")
seu <- seu[,seu$sb]
seu <- NormalizeData(seu, verbose = F) %>% ScaleData(verbose = F) %>% FindVariableFeatures(verbose = F) %>% RunPCA(npcs = 100, verbose = F)
ElbowPlot(seu, ndims = 100)
seu <- FindNeighbors(seu, dims = 1:35, verbose = F) %>% FindClusters(verbose = F) %>% RunUMAP(dims = 1:35, n.epochs = 500, verbose = F)
DimPlot(seu)
DefaultAssay(seu) <- "RNA"
FeaturePlot_scCustom(seu, "CD33")
```


Now let's see how they look after some magic.

```{r}
seu <- seurat_magic(seu, alpha = 1)
DefaultAssay(seu) <- "MAGIC"
FeaturePlot_scCustom(seu, "CD33")

```

## Changing parameters

An alpha of 1 might be too much.  We can dial it back with an alpha of 0.3 and even further by dropping t to 2.
```{r}

seu <- seurat_magic(seu, alpha = 0.3)

DefaultAssay(seu) <- "MAGIC"
FeaturePlot_scCustom(seu, "CD33")

seu <- seurat_magic(seu, alpha = 0.3, t = 2)

DefaultAssay(seu) <- "MAGIC"
FeaturePlot_scCustom(seu, "CD33")


```


