---
title: "How to perform sequence alignment"
output: html_document
date: "2024-06-04"
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

## Installing Rust

First you need to have an updated Rust installation. Go to this [site](https://www.rust-lang.org/tools/install) to learn how to install Rust.


## Installing rustytools

You will need to have the devtools package installed...

```{r, eval=F}
devtools::install_github("furlan-lab/rustytools", force=T)
```


## Running a tool to perform sequence alignment

Loading libraries
```{r, dpi=300, fig.height=4, fig.width = 6}
rm(list=ls())
suppressPackageStartupMessages({
  library(rustytools)
})

```

## Fast pairwise alignment

Using rust we have implemented an excellent pairwise alignment algorith from the rust bio package written by Johannes Köster, Vadim Nazarov, and Patrick Marks.  We can quickly perform a pairwise alignment from all the VDJ-B sequences from 10X with the adaptive clonoseq values obtained from a patient with suspected relapse of B-ALL.  The align function allows use to score the sequence similarity.  Using this approach we can easily identify the leukemic clones in the UMAP embedding.

We also see that these cells do not express CDKN2A unlike the healthy lymphod precursors, consistent with the clinical results.

```{r}
library(Seurat)
library(scCustomize)
seu<-readRDS("/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchinsonCancerCenter/Furlan_Lab - General/experiments/patient_marrows/ALL_5336/cds/240416_cds.RDS")
if(grepl("^gizmo", Sys.info()["nodename"])){
  ROOT_DIR2<-"/fh/fast/furlan_s/grp/data/ddata/BM_data"
} else {
  ROOT_DIR2<-"/Users/sfurlan/Library/CloudStorage/OneDrive-SharedLibraries-FredHutchinsonCancerCenter/Furlan_Lab - General/datasets/Healthy_BM_greenleaf"
}

#reference dataset
seur<-readRDS(file.path(ROOT_DIR2, "230329_rnaAugmented_seurat.RDS"))


DimPlot(seu, group.by = "vmR_pred", cols = seur@misc$colors)
#DimPlot(seu, group.by = "seurat_clusters", cols= as.character(pals::polychrome()))

seu$leuk_cdr3_nt<-NA
seu$leuk_cdr3_nt[seu$seurat_clusters %in% c(11,13)]<-seu$cdr3_nt[seu$seurat_clusters %in% c(11,13)]
DimPlot(seu, group.by = "leuk_cdr3_nt")+NoLegend()
putative<-names(table(seu$leuk_cdr3_nt)[order(-table(seu$leuk_cdr3_nt))])

#from adaptive clonoseq
cloneC<-"CAGGAACACCTCCATAAGCACAGCCTACATGGAGCTGAGCAGCCTGAGATCTGAGGACACGGCCGTGTATTACTGTGCGAGAGGCCTAACCCACACCCACCCCCTACTTATTGTAGTAGTACCAGCTGCTATGACTACTGGGGCCAGGGAACC"
cloneB<-"CGCGGACAAATCCACGAGCACAGCCTACATGGAGCTGAGCAGCCTGAGATCTGAGGACACGGCCGTGTATTACTGTGCGCGGCGGACTCCGTATTACTATGGTTCGGGGAGTTATACTACTACGGTATGGACGTCTGGGGCCAAGGGACC"
cloneD <- "GATGTTGGGGTTTATTACTGCATGCAAGGTACACACTGGCCCAACCTAGTGGCAGCCCAGGG"
cloneE <- "CTGATTATTACTGTGAGACCGGGACCAAGC"


seu$cloneC_score<-sapply(seu$cdr3_nt, function(seq) align(seq, cloneC, atype = "local", verbose = F))
seu$cloneB_score<-sapply(seu$cdr3_nt, function(seq) align(seq, cloneB, atype = "local", verbose = F))
seu$cloneD_score<-sapply(seu$cdr3_nt, function(seq) align(seq, cloneD, atype = "local", verbose = F))
seu$cloneE_score<-sapply(seu$cdr3_nt, function(seq) align(seq, cloneE, atype = "local", verbose = F))

FeaturePlot(seu, features = c("cloneB_score", "cloneC_score", "cloneD_score", "cloneE_score"), order = T, keep.scale="all", cols = c("grey90", "red"), min.cutoff = "q20")

Biostrings::pairwiseAlignment(putative[1], cloneC, type="global-local")
Biostrings::pairwiseAlignment(putative[2], cloneB, type="global-local")

FeaturePlot_scCustom(seu, features = "CDKN2A")
```

## Appendix
```{r Appendix}
sessionInfo()
```
